---
title: "WikipediaParseNameCatalogs"
author: "Alejandro Reyes"
date: "12/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Creating catalogs of masculine and feminine names

I generated a catalog of typically masculine and feminine names. Since Wikipedia has catalogs of masculine names and feminine names in several languages, I wrote a function to scrape these data. I ran this function for the Wikipedia pages of masculine and feminine names in Spanish, English, Italian, Portughese, French and German.

```{r}
library(rvest)


parseWikiCatalogEn <- function( wikiPage, ex1="página", ex2="anterior", nextPageLab="página siguiente", pagePre="https://es.wikipedia.org"){
  mascNames <- c()
  keepFlag <- 1
  while( keepFlag > 0 ){
    pageNames <- wikiPage %>% 
      read_html %>% 
      html_node('body #content #bodyContent #mw-content-text #mw-pages') %>% 
      html_text
    if( grepl("previous|anterior", pageNames) ){
      pageNames <- strsplit( pageNames, sprintf("\\(%s|%s\\)", ex1, ex2))[[1]][4]
    }else{
      pageNames <- strsplit(pageNames, "more\\)\\.\n")[[1]][2]
    } ### fix
    pageNames <- strsplit( pageNames, "\n" )[[1]]
    pageNames <- gsub("[A-Z]{1}$", "", pageNames)
    pageNames <- gsub("siguiente\\)", "", pageNames)
    mascNames <- c(mascNames, pageNames)
    Sys.sleep(2)
    posNodes <- wikiPage %>% 
      read_html %>% 
      html_nodes('body #content #bodyContent #mw-content-text #mw-pages > a')
    posText <- posNodes %>% html_text
    posUrl <- posNodes %>% html_attr("href")
    nextLink <- which(posText == nextPageLab )
    if( length(nextLink) > 0 ){
      keepFlag <- 1
      }else{
      keepFlag <- 0
      }
    wikiPage <- unique(posUrl[nextLink])
    wikiPage <- paste0( pagePre, wikiPage )
    Sys.sleep(2)
  }
  mascNames <- mascNames[!(mascNames == " "|mascNames == "")]
  mascNames
}

maleWikiLink <- 
  "https://es.wikipedia.org/w/index.php?title=Categor%C3%ADa:Nombres_masculinos&pageuntil=Columbano+%28nombre%29#mw-pages"
femaleWikiLink <- 
  "https://es.wikipedia.org/w/index.php?title=Categor%C3%ADa:Nombres_femeninos&pageuntil=Idoya#mw-pages"
maleNames <- parseWikiCatalogEn( maleWikiLink )
femaleNames <- parseWikiCatalogEn( femaleWikiLink )
femaleNamesExt <- lapply(
  c( italian=
       "https://en.wikipedia.org/w/index.php?title=Category:Italian_feminine_given_names&#mw-pages",
     portuguese=
       "https://en.wikipedia.org/w/index.php?title=Category:Portuguese_feminine_given_names&pageuntil=Ida+%28given+name%29#mw-pages",
    german=
      "https://en.wikipedia.org/w/index.php?title=Category:German_feminine_given_names&#mw-pages", 
    english=
      "https://en.wikipedia.org/w/index.php?title=Category:English_feminine_given_names&pageuntil=Ida+%28given+name%29#mw-pages",
    french=
      "https://en.wikipedia.org/w/index.php?title=Category:French_feminine_given_names&pageuntil=Ida+%28given+name%29#mw-pages"), 
  parseWikiCatalogEn, ex1="previous", ex2="page", nextPageLab="next page", pagePre="https://en.wikipedia.org" )
maleNamesExt <- lapply(
  c( 
    italian=
      "https://en.wikipedia.org/w/index.php?title=Category:Italian_masculine_given_names&#mw-pages",
    portuguese=
      "https://en.wikipedia.org/w/index.php?title=Category:Portuguese_masculine_given_names&pageuntil=Ida+%28given+name%29#mw-pages",
    german=
      "https://en.wikipedia.org/w/index.php?title=Category:German_masculine_given_names&#mw-pages", 
    english=
      "https://en.wikipedia.org/w/index.php?title=Category:English_masculine_given_names&pageuntil=Ida+%28given+name%29#mw-pages",
    french=
      "https://en.wikipedia.org/w/index.php?title=Category:French_masculine_given_names&pageuntil=Ida+%28given+name%29#mw-pages"), 
  parseWikiCatalogEn, ex1="previous", ex2="page", nextPageLab="next page", pagePre="https://en.wikipedia.org" )

maleNamesExt <- unlist(maleNamesExt)
femaleNamesExt <- unlist(femaleNamesExt)
femaleNamesExt <- femaleNamesExt[!femaleNamesExt %in% maleNames]
maleNamesExt <- maleNamesExt[!maleNamesExt %in% femaleNames]
maleNames <- c( maleNames, (maleNamesExt) )
femaleNames <-c( femaleNames, (femaleNamesExt) )

```

I noticed that Wikipedia catalogs were missing many names that are common in Mexico, so I added manually all missing names that appeared at least 5 times in the list of UNAM's professors. 

```{r}

maleNames <- c( maleNames, 
                c("VLADIMIR", "JESUS", "CONRAD", "BORIS", "STALIN", "REY", "GENARO", 
                  "JUSTO", "JUSTINO", "ERIK", "ANGEL", "ALAIN", "HAZAEL", "RAMON", "ALONSO", 
                  "SERAFIN", "OTONIEL", "ISAURO", "ALAN", "HIBRAIM", "EFREN", "OCTAVIO", 
                  "MOISES", "FRANCOIS", "JUVENTINO", "ANGEL", "ESTUARDO", "VLADISLAV", 
                  "ELVIA", "RUBEN", "FREDY", "MARGARITO", "OTHON", "BALTAZAR", "HERNAN", 
                  "LEONEL", "PAVEL", "NARCISO", "EMMANUEL", "FILIBERTO", "RAMIRO", 
                  "ERICK", "RAYMUNDO", "EFRAIN", "ULISES", "RENAN", "ARCADIO") )
femaleNames <- c( femaleNames, 
                  c("GUILLERMINA", "CARMEN", "EVANGELINA", "REYNA", "GUADALUPE", "NOHEMI", 
                    "MARISELA", "ALEXA", "MARTHA", "FLOR", "KATY", "TERESITA", "SELENA", 
                    "ARCELIA", "ELVA", "ILIANA", "MINELIA", "LASTENIA", "GINA", 
                    "CITLALIT", "LIBERTAD", "MARIANELA", "EVERTINA", "LEDA", "SILVANA", 
                    "ELIETE", "YLENIA", "XOCHITLALLI", "DULCE", "LIBIA", "ADELINA", 
                    "JEANNETTE", "SOLEDAD", "MAIRA", "MAYRA", "ITZEL", "SOCORRO", 
                    "GRACIELA", "IIGA", "ELBA", "CLORINDA", "KARLA", "CLAUDETTE", "IDANIA", 
                    "NORKA", "AMERICA", "TRILCE", "FABIOLA", "OLIVAMA", "ELIA", "NALLELY", 
                    "NATTIE", "MATILDE", "YURI", "CELINA", "ESMERALDA", "EVELIA", "JAZMIN", 
                    "LIGIA", "MANUELA", "NELLY", "ROSALVA", "ARIEL", "CITLALI", "IVETTE", 
                    "LILIAN", "LUCERO", "MIRNA", "NAYELI", "NIDIA", "ORALIA", "PERLA", 
                    "ROSAURA", "ROXANA", "YURI", "ZOILA", "FILIBERTO", "ISELA", "REMEDIOS", 
                    "LUCILA", "MARISOL", "MINERVA", "YADIRA", "ZOILA", "ALEJANDRINA", 
                    "EMILIA", "NIEVES", "REGINA", "YAZMIN", "ELVIA", "LIZBETH", "ERENDIRA", 
                    "FRANCISCA", "REFUGIO", "DANIELA", "MARICELA", "IVONNE", "ANTONIETA", 
                    "MIRIAM", "LEONOR", "JOSEFINA", "LILIA") )
names( maleNames ) <- NULL
names( femaleNames ) <- NULL

namesCatalog <- rbind(
  data.frame( gender=rep( "Male", length( maleNames ) ), 
              name=maleNames, stringsAsFactors=FALSE),
  data.frame( gender=rep( "Female", length( femaleNames ) ), 
              name=femaleNames, stringsAsFactors = FALSE ) )

namesCatalog <- namesCatalog %>%
  dplyr::filter( nchar(name) > 1 ) %>%
  dplyr::mutate( nameMod=gsub("\\(.*\\).*", "", name ) ) %>%
  dplyr::mutate( nameMod=gsub(" $", "", nameMod ))

unwanted_array = list('Š'='S', 'š'='s', 'Ž'='Z', 'ž'='z', 'À'='A', 'Á'='A', 'Â'='A', 'Ã'='A', 'Ä'='A', 'Å'='A', 'Æ'='A', 'Ç'='C', 'È'='E', 'É'='E', 'Ê'='E', 'Ë'='E', 'Ì'='I', 'Í'='I', 'Î'='I', 'Ï'='I', 'Ñ'='N', 'Ò'='O', 'Ó'='O', 'Ô'='O', 'Õ'='O', 'Ö'='O', 'Ø'='O', 'Ù'='U', 'Ú'='U', 'Û'='U', 'Ü'='U', 'Ý'='Y', 'Þ'='B', 'ß'='Ss', 'à'='a', 'á'='a', 'â'='a', 'ã'='a', 'ä'='a', 'å'='a', 'æ'='a', 'ā'='a', 'ç'='c', 'è'='e', 'é'='e', 'ê'='e', 'ë'='e', 'ì'='i', 'í'='i', 'ī'='i', 'î'='i', 'ï'='i', 'ł'='l', 'ð'='o', 'ñ'='n', 'ò'='o', 'ó'='o', 'ō'='o', 'ô'='o', 'õ'='o', 'ö'='o', 'ø'='o', 'ś'='s', 'ù'='u', 'ú'='u', 'û'='u', 'ū'='u', 'ý'='y', 'ý'='y', 'þ'='b', 'ÿ'='y', 'ż'='z' )

namesCatalog$nameMod <- chartr( paste(names(unwanted_array), collapse=''),
         paste(unwanted_array, collapse=''),
         namesCatalog$nameMod )
namesCatalog$nameMod <- toupper(namesCatalog$nameMod)
namesCatalog <- dplyr::rename( namesCatalog, nombre=nameMod )
namesCatalog <- unique(namesCatalog)
```

I discarded most names that are given to both males and females. But for some of these ambiguous cases, I assigned a gender based on the gender that is more common in Mexico for that name. For example, the name "Angel" is listed in Wikipedia as a name for both females and males, but it is more common for males in Mexico. 

```{r}

namesCatalog <- namesCatalog[!(namesCatalog$nombre == "ANGEL" & namesCatalog$gender == "Female"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "ELVIA" & namesCatalog$gender == "Male"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "FILIBERTO" & namesCatalog$gender == "Female"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "ISIDORO" & namesCatalog$gender == "Female"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "ARIEL" & namesCatalog$gender == "Male"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "EVELYN" & namesCatalog$gender == "Male"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "ROSARIO" & namesCatalog$gender == "Male"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "AIME" & namesCatalog$gender == "Male"),]
namesCatalog <- namesCatalog[!(namesCatalog$nombre == "SOL" & namesCatalog$gender == "Male"),]

neutralNames <- intersect( 
  namesCatalog$nombre[namesCatalog$gender == "Male"],
  namesCatalog$nombre[namesCatalog$gender == "Female"] )

namesCatalog <- namesCatalog[!namesCatalog$nombre %in% neutralNames,]
```

I collected `r format( sum(namesCatalog$gender == "Male"), big.mark=",")` male names and `r format(sum(namesCatalog$gender == "Female"), big.mark=",")` female names. 

